Typed Lua: An Optional Type System for Lua

### 2014-03-16 doc/paper/typedlua.tex
fixing small typos, and giving more references to related work

Dynamically typed languages such as Lua avoid static types in favor of
run-time {\em type tags} that classify the values they compute, and
their implementations use these tags to perform run-time (or dynamic)
checking and guarantee that only valid operations are
performed~\citep{pierce2002tpl}.

@@

Ideally, programming languages should offer programmers the
option to choose between static and dynamic typing:
\textit{optional type systems}~\citep{bracha2004pluggable} and
\textit{gradual typing}~\citep{siek2006gradual} are two approaches
that offer programmers the option to use type annotations where static
typing is needed, incrementally migrating a system from dynamic
to static types. The difference between these two approaches is the
way they treat run-time semantics: while optional type systems
do not affect the run-time semantics,
gradual typing uses run-time checks to ensure that dynamically typed
code does not violate the invariants of statically typed code.

@@

Typed Lua allows statically typed Lua code to coexist and
interact with  ynamically typed code. The Typed Lua compiler
warns the programmer about type errors, but always generates
Lua code that runs in unmodified Lua implementations. The
programmer can enjoy some of the benefits of static types even
without converting existing Lua modules to typed Lua:
a dynamically typed module can export a statically typed
interface, and statically typed users of the module will have
their use of the module checked by the compiler.

Unlike gradual type systems, Typed Lua does not insert run-time
checks between dynamically and statically typed parts of the
program. Unlike some optional type systems, the statically
typed subset of Typed Lua is sound by design, so a later
version of the Typed Lua compiler can swicth to gradual
instead of just optional typing.

@@

The {\em dynamic type} $\Any$ allows dynamically typed code to
interoperate with statically typed code; it is a subtype of $\Top$, but
neither a supertype nor a subtype of any other type. We relate $\Any$
to other types with the {\em consistency} and {\em consistent-subtype}
relationships used by gradual type systems. In practice, we can
pass a value of the dynamic type anytime we want a value of some
other type, and can pass any value where a value of the dynamic type
is expected, but these operations are tracked by the type system,
and the programmer can choose to be warned about them.

\begin{figure}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
\multicolumn{3}{c}{\textbf{First-class types}} \\
T ::= & \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Top & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; T \cup T & \textit{disjoint union types}\\
& | \; S \rightarrow S & \textit{function types}\\
& | \; \{F, ..., F\} & \textit{table types}\\
& | \; X & \textit{type variables}\\
& | \; \mu X.T & \textit{recursive types}\\
& | \; X_i & \textit{projection types}\\
L ::= & \multicolumn{2}{l}{\False \; | \; \True \; | \; {<}{\it number}{>} \; | \; {<}{\it string}{>} } \\
B ::= & \multicolumn{2}{l}{\Nil \; | \; \Boolean \; | \; \Number \; | \; \String} \\
F ::= & T:T \; | \; \Const \; T:T & \textit{field types} \\
\multicolumn{3}{c}{} \\
\multicolumn{3}{c}{\textbf{Second-class types}} \\
S ::= &  \multicolumn{2}{l}{\Void \; |\; V \; | \; S \cup S} \\
V ::= & \;\; T & \\
& | \; T* & \textit{vararg types}\\
& | \; T \times V & \textit{tuple types}
\end{array}
$$
\dend
\caption{Abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure}

Typed Lua allows optional type annotations in variable function
declarations. It assigns the dynamic type to any parameter that
does not have an type annotation, but assigns more precise
types to unnanotated variables, based on the type of the
expression that gives the initial value of the variable.

@@

As we saw in Figure \ref{fig:typelang}, a second-class type in
Typed Lua can be the type $\Void$ (the empty tuple),
a tuple of first-class types optionally ending in a variadic type,
or a union of these tuples. A variadic type $T*$ is a generator for a
sequence of values of type $T \cup \Nil$. Unions of tuples play
an important part in functions that are overloaded on the return type,
together with {\em projection types}. Both are explained in the next
section.

@@

In this example, the second parameter is optional but, in the first
line of the function, we declare a new variable that is guaranteed to
have type $\String$ instead of $\String \cup \Nil$. In Lua, any value
except {\tt nil} and {\tt false} are "truthy", so the short-circuiting
{\tt or} operator is a common way of giving a default value to an
optional parameter. Typed Lua encodes this idiom with a typing rule: if
the left side of {\tt or} has type $T \cup Nil$ and the right side
has type $T$ then the {\tt or} expression has type $T$.

@@

Typed Lua lets assignment change the type of a local variable in
cases that respect a simple heuristic: the new type must be a
subtype of the previous one, and the variable must be local
to the current function. The new type only applies for the
remainder of the current scope.

@@

There is also special syntax for this idiom: we could
annotate the return type of {\tt idiv} with {\tt (number, number)?}
to denote the same union\footnote{The parentheses are always
necessary here: {\tt number?} is {\tt number|nil}, while {\tt (number)}
is {\tt (number)|(nil, string)}.}.

@@

If we need to check a projection type
against some other type, we take the union of the corresponding
component in each tuple. But if we a variable with a projection
type appears in a type predicate, the predicate discriminates
against all tuples in the union. In the example above,
$X$ is $(\Number \times \Number \times \Nil *) \cup
(\Nil \times \String \times \Nil *)$ outside of the {\tt if}
statement, but $\Number \times \Number \times \Nil *$ in the
{\tt then} block and $\Nil \times \String \times \Nil *$ in
the {\tt else} block.

Notice that we could also discriminate using {\tt r}, using
{\tt type(r) == "number"} as our predicate, with the same
result. The first form is more succint, and more idiomatic.
We can also use projection types to write overloaded functions
where the type of a parameter depends on the type of another
parameter.

@@

Typed Lua tags a variable with a table type as either
{\em open} or {\em closed}. If a variable gets its type
from a table constructor then it is open, otherwise it is
always closed. The type of an open variable may change
by field assignment, subject to two restrictions: the
variable must be local to the current block, and the new type
must be a subtype of the old type.

@@

Lua's built-in support for object oriented programming is
minimal. The basic mechanism is the {\tt :} syntactic
sugar for method calls and method declarations.
The Lua compiler translates {\tt obj:method(args)}
to an operation that evaluates {\tt obj}, looks
for a field named ``method" in the result, then calls
it with the result of evaluating {\tt obj} as the
first argument, followed by the result of evaluating
the argument list in the original expression.

@@

\begin{verbatim}
    local Shape = { x = 0, y = 0 }
    const function Shape:new(x: number, y: number)
      local s = setmetatable({},
                             { __index = self })
      s.x = x
      s.y = y
      return shape
    end
    const function Shape:move(dx: number,
                              dy: number)
      self.x = self.x + dx
      self.y = self.y + dy
    end
    local shape1 = Shape:new(0, 5)
    local shape2: Shape = Shape:new(10, 10)
\end{verbatim}

@@

\begin{verbatim}
    local Circle = setmetatable({},
                                { __index = Shape })
    Circle.radius = 0
    const function Circle:new(x: number,
                              y: number,
                              radius: number)
      local c = setmetatable(Shape:new(x, y),
                             { __index = self })
      c.radius = radius
      return c
    end
    const function Circle:area()
      return math.pi * self.radius * self.radius
    end
\end{verbatim}

In the first line of the redefinition of {\tt new},
notice how we can call {\tt Shape}'s constructor inside
the overriden constructor. A limitation of this class
system is that the overriden constructors must be a
subtype of the original constructor.

@@

\textit{Dynamic typing}~\citep{henglein1994dts} is
another approach for optimizing dynamically-typed programs.
First the program is translated to a program that uses
a {\tt Dynamic} type and explicit coercions and runtime checks,
then a static analysis removes some  of these coercions and
checks.

@@

Tidal Lock \citep{tidallock} is a prototype of another optional
type system for Lua. Compared with Typed Lua, Tidal Lock
has richer record types, and a more robust system of
incremental evolution of these record types, but it lacks
unions and support for expressing objects and classes.

>>>>

Dynamically typed languages such as Lua avoid static types in favor of
runtime {\em type tags} that classify the values they compute, and
their implementations use these tags to perform runtime (or dynamic)
checking and guarantee that only valid operations are
performed~\citep{pierce2002tpl}.

@@

Ideally, programming languages should offer programmers the
option to choose between static and dynamic typing:
\textit{optional type systems}~\citep{bracha2004pluggable} and
\textit{gradual typing}~\citep{siek2006gradual} are two approaches
that offer programmers the option to use type annotations where static
typing is needed, incrementally migrating a system from dynamic
to static types. The difference between these two approaches is the
way they treat runtime semantics: while optional type systems
do not affect the runtime semantics,
gradual typing uses runtime checks to ensure that dynamically typed
code does not violate the invariants of statically typed code.

@@

Typed Lua allows statically typed Lua code to coexist and
interact with dynamically typed code. The Typed Lua compiler
warns the programmer about type errors, but always generates
Lua code that runs in unmodified Lua implementations. The
programmer can enjoy some of the benefits of static types even
without converting existing Lua modules to Typed Lua:
a dynamically typed module can export a statically typed
interface, and statically typed users of the module will have
their use of the module checked by the compiler.

Unlike gradual type systems, Typed Lua does not insert runtime
checks between dynamically and statically typed parts of the
program. Unlike some optional type systems, the statically
typed subset of Typed Lua is sound by design, so a later
version of the Typed Lua compiler can swicth to gradual
instead of just optional typing.

@@

The {\em dynamic type} $\Any$ allows dynamically typed code to
interoperate with statically typed code; it is a subtype of 
$\Top$, but
neither a supertype nor a subtype of any other type. We relate 
$\Any$
to other types with the {\em consistency} and {\em consistent-
subtype}
relationships used by gradual type systems~\citep{siek2006gradual,siek2007objects}.
 In practice, we can
pass a value of the dynamic type anytime we want a value of some
other type, and can pass any value where a value of the dynamic type
is expected, but these operations are tracked by the type system,
and the programmer can choose to be warned about them.

\begin{figure}[!ht]
\textbf{Type Language}\\
\dstart
$$
\begin{array}{rlr}
\multicolumn{3}{c}{\textbf{First-class types}} \\
T ::= & \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Top & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; T \cup T & \textit{disjoint union types}\\
& | \; S \rightarrow S & \textit{function types}\\
& | \; \{F, ..., F\} & \textit{table types}\\
& | \; X & \textit{type variables}\\
& | \; \mu X.T & \textit{recursive types}\\
& | \; X_i & \textit{projection types}\\
L ::= & \multicolumn{2}{l}{\False \; | \; \True \; | \; {<}{\it number}{>} \; | \; {<}{\it string}{>} } \\
B ::= & \multicolumn{2}{l}{\Nil \; | \; \Boolean \; | \; \Number \; | \; \String} \\
F ::= & T:T \; | \; \Const \; T:T & \textit{field types} \\
\multicolumn{3}{c}{} \\
\multicolumn{3}{c}{\textbf{Second-class types}} \\
S ::= &  \multicolumn{2}{l}{\; V \; | \; V \cup S} \\
V ::= & \;\; T & \\
& | \; T* & \textit{vararg types}\\
& | \; T \times V & \textit{tuple types}
\end{array}
$$
\dend
\caption{Abstract syntax of Typed Lua types}
\label{fig:typelang}
\end{figure}

Typed Lua allows optional type annotations in variable function
declarations. It assigns the dynamic type to any parameter that
does not have a type annotation, but assigns more precise
types to unnanotated variables, based on the type of the
expression that gives the initial value of the variable.

@@

As we can see in Figure \ref{fig:typelang}, a second-class
type in Typed Lua can be a tuple of first-class types
optionally ending in a variadic type, or a union of these
tuples. A variadic type $T*$ is a generator for a
sequence of values of type $T \cup \Nil$. Unions of tuples play
an important part in functions that are overloaded on the
return type, together with {\em projection types}. Both are
explained in the next
section.

@@

In this example, the second parameter is optional but, in the
first line of the function, we declare a new variable that is
guaranteed to have type $\String$ instead of $\String \cup \Nil$.
In Lua, any value except {\tt nil} and {\tt false} are ``truthy",
so the short-circuiting {\tt or} operator is a common way of
giving a default value to an optional parameter. Typed Lua
encodes this idiom with a typing rule: if
the left side of {\tt or} has type $T \cup \Nil$ and the right side has type $T$ then the {\tt or} expression has type $T$.

@@

Typed Lua lets assignment change the type of a local variable in
cases that respect a simple heuristic: the new type must be a
subtype of the previous one, the variable must be local
to the current function, and the variable must not have been
assigned to in another function. The new type only applies for
the remainder of the current scope. In the case of {\tt
greeting}, the assignment on line three changes its type to
$\String$.

@@

There is also special syntax for this idiom: we could
annotate the return type of {\tt idiv} with {\tt (number, number)?}
to denote the same union\footnote{The parentheses are always
necessary here: {\tt number?} is {\tt number|nil}, while {\tt (number)?}
is {\tt (number)|(nil, string)}.}.

@@

If we need to check a projection type
against some other type, we take the union of the corresponding
component in each tuple. But if a variable with a projection
type appears in a type predicate, the predicate discriminates
against all tuples in the union. In the example above,
$X$ is $(\Number \times \Number \times \Nil *) \cup
(\Nil \times \String \times \Nil *)$ outside of the {\tt if}
statement, but $\Number \times \Number \times \Nil *$ in the
{\tt then} block and $\Nil \times \String \times \Nil *$ in
the {\tt else} block.

Notice that we could also discriminate {\tt r} using
{\tt type(r) == "number"} as our predicate, with the same
result. The first form is more succint, and more idiomatic.
We can also use projection types to write overloaded functions
where the type of a parameter depends on the type of another
parameter.

@@

Typed Lua tags a variable with a table type as either
{\em open} or {\em closed}. If a variable gets its type
from a table constructor then it is open, otherwise it is
always closed. The type of an open variable may change
by field assignment, subject to two restrictions: the
variable must be local to the current block, the new type
must be a subtype of the old type, and the variable
cannot have been assigned to in another function.

@@

Lua's built-in support for object-oriented programming is
minimal. The basic mechanism is the {\tt :} syntactic
sugar for method calls and method declarations.
The Lua compiler translates {\tt obj:method(args)}
to an operation that evaluates {\tt obj}, looks
for a field named ``method" in the result, then calls
it with the result of evaluating {\tt obj} as the
first argument, followed by the result of evaluating
the argument list in the original expression.

@@

\begin{verbatim}
    local Shape = { x = 0, y = 0 }
    const function Shape:new(x: number, y: number)
      local s = setmetatable({},
                             { __index = self })
      s.x = x
      s.y = y
      return s
    end
    const function Shape:move(dx: number,
                              dy: number)
      self.x = self.x + dx
      self.y = self.y + dy
    end
    local shape1 = Shape:new(0, 5)
    local shape2: Shape = Shape:new(10, 10)
\end{verbatim}

@@

\begin{verbatim}
    local Circle = setmetatable({},
                                { __index = Shape })
    Circle.radius = 0
    const function Circle:new(x: number,
                              y: number,
                              radius: value)
      local c = setmetatable(Shape:new(x, y),
                             { __index = self })
      c.radius = tonumber(radius)
      return c
    end
    const function Circle:area()
      return math.pi * self.radius * self.radius
    end
\end{verbatim}

In the first line of the redefinition of {\tt new},
notice how we can call {\tt Shape}'s constructor inside
the overriden constructor. A limitation of this class
system is that the overriden constructor must be a
subtype of the original constructor, so the type of
{\tt radius} has to be very permissive.

@@

While the goal of soft typing is catching errors,
\textit{dynamic typing}~\citep{henglein1994dts} is
another approach for optimizing dynamically-typed programs.
First the program is translated to a program that uses
a {\tt Dynamic} type and explicit coercions and runtime checks,
then a static analysis removes some  of these coercions and
checks.

@@

Grace~\citet{black2013sg} is an object-oriented language
with optional typing. Grace is not an dynamically typed
language that has been extended with an optional type system,
but a language that has been designed from the ground up
to have both static and dynamic typing. \citet{homer2013modules}
explores some useful patterns that derive from Grace's
use of objects as modules and its brand of optional
structural typing, which can also be expressed with
Typed Lua's modules as tables.

The main feature of Typed Racket's type system is
{\em ocurrence typing}~\citep{tobin:occur}, where
the type of a variable can be refined through tests
using predicates that check the runtime type of
the variable. As these types of checks are common in
other languages, related systems have
appeared~\citep{guha:flow,winther:flow,pearce:whiley}.

Tidal Lock~\citep{tidallock} is a prototype of another optional
type system for Lua. Compared with Typed Lua, Tidal Lock
has richer record types, and a more robust system of
incremental evolution of these record types, but it lacks
unions and support for expressing objects and classes.
